A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-08-07 doc/large_strain.tex
Update tables

\begin{table}
  \centering
  \resizebox{\textwidth}{!}{
  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
  \hline
                & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\
  \hline
  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
  \hline
  2& \pgfmathprintnumber{0.71685} & \pgfmathprintnumber{0.8747712} & \pgfmathprintnumber{0.07427} & \pgfmathprintnumber{8.57318} & \pgfmathprintnumber{9.65194560388} & \pgfmathprintnumber{0.3939} & \pgfmathprintnumber{1.6528706} & \pgfmathprintnumber{1.81987814166} & \pgfmathprintnumber{0.04142} & \pgfmathprintnumber{16.0073657} & \pgfmathprintnumber{17.306856591} \\
  4& \pgfmathprintnumber{0.47603} & \pgfmathprintnumber{1.3088965} & \pgfmathprintnumber{0.0511} & \pgfmathprintnumber{12.3860771} & \pgfmathprintnumber{9.3156555773} & \pgfmathprintnumber{0.25747} & \pgfmathprintnumber{2.3690572} & \pgfmathprintnumber{1.84887559716} & \pgfmathprintnumber{0.02786} & \pgfmathprintnumber{22.2420699} & \pgfmathprintnumber{17.0865039483} \\
  6& \pgfmathprintnumber{0.24683} & \pgfmathprintnumber{1.5497601} & \pgfmathprintnumber{0.02636} & \pgfmathprintnumber{14.8125524} & \pgfmathprintnumber{9.36380880121} & \pgfmathprintnumber{0.13089} & \pgfmathprintnumber{2.8544567} & \pgfmathprintnumber{1.88578195431} & \pgfmathprintnumber{0.01487} & \pgfmathprintnumber{26.0509717} & \pgfmathprintnumber{16.5991930061} \\
  \hline
  \end{tabular}
  }
  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
  vectorization and parallelization.}
  \label{tab:numbers_2d}
\end{table}

\begin{table}
  \centering
  \resizebox{\textwidth}{!}{
  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
  \hline
                & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\
  \hline
  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
  \hline
  2& \pgfmathprintnumber{1.82927} & \pgfmathprintnumber{1.4408182} & \pgfmathprintnumber{0.19184} & \pgfmathprintnumber{13.75528} & \pgfmathprintnumber{9.5353940784} & \pgfmathprintnumber{1.24738} & \pgfmathprintnumber{2.2390063} & \pgfmathprintnumber{1.46648976254} & \pgfmathprintnumber{0.13397} & \pgfmathprintnumber{21.0187594} & \pgfmathprintnumber{13.6543255953} \\
  4& \pgfmathprintnumber{1.14988} & \pgfmathprintnumber{1.7483867} & \pgfmathprintnumber{0.1211} & \pgfmathprintnumber{16.6162812} & \pgfmathprintnumber{9.49529314616} & \pgfmathprintnumber{0.74974} & \pgfmathprintnumber{2.7527409} & \pgfmathprintnumber{1.53370501774} & \pgfmathprintnumber{0.08166} & \pgfmathprintnumber{25.4865287} & \pgfmathprintnumber{14.0813127602} \\
  \hline
  \end{tabular}
  }
  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 3D for various combinations of polynomial degrees,
  vectorization and parallelization.}
  \label{tab:numbers_3d}
\end{table}

Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI. For these rather small count of MPI processes, the results confirm that the performance gained is independent. Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
\todo[inline]{Do we want to update Tables \ref{tab:numbers_2d} and \ref{tab:numbers_3d}?}

@@

An extension of this study from the node level to the cluster level is beyond the scope of this work.
However, we have performed a preliminary study of the weak scalability of the implementation.
Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes\footnote{%
This corresponds to 64 nodes, which is the maximum we could request on the RRZE Emmy HPC cluster.
} of the wall time per matrix-vector product and per iteration in the CG solver for the 3D problem with a tri-quadratic basis.
This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
\todo[inline]{Do we want to update? I can run on up to 8 nodes with CSL. This allows me to do 3 steps of weak scaling in 2D, 1 socket, 2 nodes, 8 nodes, and two steps in 3D, 1 node and 8 nodes. Not very large but enough to show our point?}

\begin{table}
  \centering
  \begin{tabular}{|r|r|c|r|c|}
  \hline
  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
  \hline
  20 & 4,442,880 & 3 & \pgfmathprintnumber{0.0675} & \pgfmathprintnumber{0.529553679131} \\
  160 & 35,071,488 & 4 & \pgfmathprintnumber{0.0705} & \pgfmathprintnumber{0.562637362637} \\
  1280 & 278,694,912 & 5 & \pgfmathprintnumber{0.083} & \pgfmathprintnumber{0.688792165397} \\
  \hline
  \end{tabular}
  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor4} in 3D for quadratic polynomial basis.}
  \label{tab:weak_3d}
\end{table}

>>>>

\begin{table}
  \caption{Wall-clock time in seconds and performance in GFlops of Algorithms \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
  vectorization and parallelization on Cascade Lake.}
  \label{tab:numbers_2d}
  \centering
  \resizebox{\textwidth}{!}{
  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
  \hline
                & \multicolumn{2}{c|}{auto-vect serial} & \multicolumn{3}{c|}{auto-vect MPI 20 cores} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{SIMD+MPI 20 cores}  \\
                & \multicolumn{2}{c|}{3.8 GHz} & \multicolumn{3}{c|}{2.4 GHz} & \multicolumn{3}{c|}{3.7 GHz} & \multicolumn{3}{c|}{2.0 GHz} \\
  \hline
  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
  \hline
& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor2}}\\
\hline
2& \pgfmathprintnumber{0.67518} & \pgfmathprintnumber{2.1798691} & \pgfmathprintnumber{0.04811} & \pgfmathprintnumber{31.270185} & \pgfmathprintnumber{14.0340885471} & \pgfmathprintnumber{0.21837} & \pgfmathprintnumber{6.6211982} & \pgfmathprintnumber{3.09190822915} & \pgfmathprintnumber{0.02194} & \pgfmathprintnumber{67.8161088} & \pgfmathprintnumber{30.773928897} \\
4& \pgfmathprintnumber{0.35754} & \pgfmathprintnumber{3.6114434} & \pgfmathprintnumber{0.02991} & \pgfmathprintnumber{43.3813195} & \pgfmathprintnumber{11.9538615848} & \pgfmathprintnumber{0.11568} & \pgfmathprintnumber{11.0068739} & \pgfmathprintnumber{3.09076763485} & \pgfmathprintnumber{0.01438} & \pgfmathprintnumber{89.247277} & \pgfmathprintnumber{24.8636995828} \\
6& \pgfmathprintnumber{0.17724} & \pgfmathprintnumber{4.4013124} & \pgfmathprintnumber{0.01489} & \pgfmathprintnumber{52.73939} & \pgfmathprintnumber{11.9032907992} & \pgfmathprintnumber{0.05934} & \pgfmathprintnumber{12.9981312} & \pgfmathprintnumber{2.9868554095} & \pgfmathprintnumber{0.00732} & \pgfmathprintnumber{106.9726898} & \pgfmathprintnumber{24.2131147541} \\
\hline
& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor4}}\\
\hline
2& \pgfmathprintnumber{0.62536} & \pgfmathprintnumber{2.623296} & \pgfmathprintnumber{0.04582} & \pgfmathprintnumber{36.1519843} & \pgfmathprintnumber{13.6481885639} & \pgfmathprintnumber{0.22937} & \pgfmathprintnumber{6.925918} & \pgfmathprintnumber{2.72642455421} & \pgfmathprintnumber{0.02532} & \pgfmathprintnumber{63.6564648} & \pgfmathprintnumber{24.6982622433} \\
4& \pgfmathprintnumber{0.36549} & \pgfmathprintnumber{3.8534134} & \pgfmathprintnumber{0.03113} & \pgfmathprintnumber{45.7061894} & \pgfmathprintnumber{11.7407645358} & \pgfmathprintnumber{0.14346} & \pgfmathprintnumber{9.5663639} & \pgfmathprintnumber{2.54767879548} & \pgfmathprintnumber{0.01862} & \pgfmathprintnumber{74.2778413} & \pgfmathprintnumber{19.6288936627} \\
6& \pgfmathprintnumber{0.18374} & \pgfmathprintnumber{4.5579733} & \pgfmathprintnumber{0.01572} & \pgfmathprintnumber{54.0989777} & \pgfmathprintnumber{11.6882951654} & \pgfmathprintnumber{0.07388} & \pgfmathprintnumber{11.097412} & \pgfmathprintnumber{2.4870059556} & \pgfmathprintnumber{0.00946} & \pgfmathprintnumber{88.2340296} & \pgfmathprintnumber{19.422832981} \\
  \hline
  \end{tabular}
  }
\end{table}

\begin{table}
  \caption{Wall-clock time in seconds and performance in GFlops of Algorithms~\ref{alg:mf_tensor2} and \ref{alg:mf_tensor4} in 3D for various combinations of polynomial degrees,
  vectorization and parallelization on Cascade Lake.}
  \centering
  \resizebox{\textwidth}{!}{
  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
  \hline
                & \multicolumn{2}{c|}{auto-vect serial} & \multicolumn{3}{c|}{auto-vect MPI 20 cores} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{SIMD+MPI 20 cores}  \\
                & \multicolumn{2}{c|}{3.8 GHz} & \multicolumn{3}{c|}{2.4 GHz} & \multicolumn{3}{c|}{3.7 GHz} & \multicolumn{3}{c|}{2.0 GHz} \\
  \hline
  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
  \hline
& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor2}}\\
\hline
2& \pgfmathprintnumber{0.33065} & \pgfmathprintnumber{4.341083} & \pgfmathprintnumber{0.02814} & \pgfmathprintnumber{51.7484607} & \pgfmathprintnumber{11.750177683} & \pgfmathprintnumber{0.12258} & \pgfmathprintnumber{11.7249488} & \pgfmathprintnumber{2.6974220917} & \pgfmathprintnumber{0.01317} & \pgfmathprintnumber{112.4745656} & \pgfmathprintnumber{25.106302202} \\
4& \pgfmathprintnumber{0.18685} & \pgfmathprintnumber{5.461242} & \pgfmathprintnumber{0.01589} & \pgfmathprintnumber{64.7662171} & \pgfmathprintnumber{11.7589679043} & \pgfmathprintnumber{0.0695} & \pgfmathprintnumber{14.7126397} & \pgfmathprintnumber{2.68848920863} & \pgfmathprintnumber{0.00803} & \pgfmathprintnumber{129.6590805} & \pgfmathprintnumber{23.2689912827} \\
  \hline
& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor4}}\\
  \hline
2& \pgfmathprintnumber{0.38409} & \pgfmathprintnumber{4.4971589} & \pgfmathprintnumber{0.03472} & \pgfmathprintnumber{51.9008095} & \pgfmathprintnumber{11.0625} & \pgfmathprintnumber{0.22335} & \pgfmathprintnumber{7.7420018} & \pgfmathprintnumber{1.719677636} & \pgfmathprintnumber{0.0243} & \pgfmathprintnumber{72.555647} & \pgfmathprintnumber{15.8061728395} \\
4& \pgfmathprintnumber{0.22053} & \pgfmathprintnumber{5.393482} & \pgfmathprintnumber{0.01898} & \pgfmathprintnumber{63.1076193} & \pgfmathprintnumber{11.6190727081} & \pgfmathprintnumber{0.12765} & \pgfmathprintnumber{9.3340763} & \pgfmathprintnumber{1.72761457109} & \pgfmathprintnumber{0.01464} & \pgfmathprintnumber{82.3893605} & \pgfmathprintnumber{15.0635245902} \\
\hline
  \end{tabular}
  }
  \label{tab:numbers_3d}
\end{table}

Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to the MPI parallelization.
{\color{red}
When run in serial, explicit SIMD vectorization via instrinsics provides a speedup of up to $3\times$ over the auto-vectorization case. This is less than the theoretical factor of 8, but in line with previous experiments presented in \cite[Figure 6]{kronbichler2017fastcomputer} with a speedup of around $1.5-2.5\times$ with 8-wide vectorization. The reason for sub-optimal speedup is mostly related to the memory access in the gather/scatter operations as well as the limited memory bandwidth. When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effect. For the auto-vectorized case, the speedup is limited due to different turbo clock frequencies, which permit a speedup of a factor 12.6.\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic limits.} For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited. Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity, the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
}

@@

An extension of this study from the node level to the cluster level is beyond the scope of this work.
However, we have performed a preliminary study of the weak scalability of the implementation.
Tables~\ref{tab:weak_2d} \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake, of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.
This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.

\begin{table}
  \centering
  \begin{tabular}{|r|r|c|r|c|}
  \hline
  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
  \hline
  20 & 2,887,680 & 5 & \pgfmathprintnumber{0.0033900} & \pgfmathprintnumber{0.045574} \\
  80 & 11,542,528 & 6 & \pgfmathprintnumber{0.0035350} & \pgfmathprintnumber{0.047788} \\
  320 & 46,153,728 & 7 & \pgfmathprintnumber{0.0037879} & \pgfmathprintnumber{0.052012} \\
  \hline
  \end{tabular}
  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 2D for a quartic polynomial basis.}
  \label{tab:weak_2d}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|r|r|c|r|c|}
  \hline
  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
  \hline
  40 & 4,442,880 & 3 & \pgfmathprintnumber{0.0070667} & \pgfmathprintnumber{0.064328} \\
  320 & 35,071,488 & 4 & \pgfmathprintnumber{0.0075333} & \pgfmathprintnumber{0.071448} \\
  \hline
  \end{tabular}
  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 3D for a quadratic polynomial basis.}
  \label{tab:weak_3d}
\end{table}
